---
title: "Minimum Wage Project"
output:
  flexdashboard::flex_dashboard:
    theme: 
      version: 4
      bootswatch: cerulean
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(dplyr)
library(ggplot2)
```

```{r}
minwage = read.csv("myminwage.csv")
PCE = read.csv("myPCE.csv")
EMP = read.csv("myEMP.csv")
INC = read.csv("myINC.csv")
GDP = read.csv("myGDP.csv")
subPCE = read.csv("subPCE.csv")
subEMP = read.csv("subEMP.csv")
subINC = read.csv("subINC.csv")
```

State Data
===================================== 

## Column {.sidebar}
```{r}
selectInput("State", label="State", choices = sort(na.omit(minwage$GeoName)))
```

Minimum Wage Date Source: U.S. Department of Labor

Indicator Data Source: Bureau of Economic Analysis

## Column

### Minimum Wage Change 2011-2020
```{r}
renderGauge({
  x <- minwage %>% filter(GeoName == input$State) %>% select(X10change) %>% pull(1)
  gauge(x, min = 0, max = 6.75, gaugeSectors(success = c(4,6.75), warning = c(2,3.99), danger = c(0,1.99)))})
```

### PCE Per Capita Change 2011-2020
```{r}
renderGauge({
  x <- PCE %>% filter(GeoName == input$State, trimws(Description) == "Personal consumption expenditures") %>% select(pc10change) %>% pull(1)
  gauge(x, min = 0, max = 127000, gaugeSectors(success = c(0,14999), warning = c(15000,49999), danger = c(50000,127000)))})
```
### INC Per Capita Change 2011-2020
```{r}
renderGauge({
  x <- INC %>% filter(GeoName == input$State, trimws(Description) == "Per capita personal income (dollars) 2/") %>% select(X10change.x) %>% pull(1)
  gauge(x, min = 10000, max = 26000, gaugeSectors(success = c(16000,26000), warning = c(13000,16000), danger = c(10000,13000)))})
```

## Column

### EMP Per Capita Change 2011-2020
```{r}
renderGauge({
  x <- EMP %>% filter(GeoName == input$State, trimws(Description) == "Total employment (number of jobs)") %>% select(pc10change) %>% pull(1)
  gauge(x, min = -.5, max = .5, gaugeSectors(success = c(.05,.4), warning = c(-.05,.05), danger = c(-.05,-.5)))})
```

### GDP Per Capita Change 2011-2020
```{r}
renderGauge({
  x <- GDP %>% filter(GeoName == input$State) %>% select(pc10change) %>% pull(1)
  gauge(x, min = -11000, max = 32000, gaugeSectors(success = c(15000,32000), warning = c(5000,15000), danger = c(-11000,5000)))})
```


PCE
===================================== 

## Column {data-width=550}

```{r}
ggplot(subPCE, aes(x = X10change.y, y = pc10change)) + 
  geom_point() + 
  geom_smooth() + 
  xlab("Change in Minimum Wage ($)") +
  ylab("Change in Personal Consumption Expenditures Per Capita ($)") 
```

### Correlation Coefficient
```{r}
valueBox(0.2772036)
```

## Column {data-width=350}

### Personal Consumption Expenditure

* Personal Consumption Expenditure is the measure spending on goods and services. 
* PCE is a way of measuring inflation. 
* Changes in minimum wage and changes in total personal consumption expenditure are very weakly correlated. 


EMP
===================================== 

## Column {data-width=550}

```{r}
ggplot(subEMP, aes(x = X10change.y, y = pc10change)) + 
  geom_point() + 
  geom_smooth() + 
  xlab("Change in Minimum Wage ($)") +
  ylab("Change in Total Employment per Capita (Number of Jobs)")  
```

### Correlation Coefficient
```{r}
valueBox(0.1210715)
```

## Column {data-width=350}

### Employment

* Employment change over the past 10 years (per capita). 
* Changes in minimum wage do not affect changes in total employment. 

INC
===================================== 

## Column {data-width=550}

```{r}
ggplot(subINC, aes(x = X10change.y, y = X10change.x)) + 
  geom_point() + 
  geom_smooth() + 
  xlab("Change in Minimum Wage ($)") +
  ylab("Change in Personal Income per Capita ($)") 
```

### Correlation Coefficient
```{r}
valueBox(0.5732798)
```

## Column {data-width=350}

### Personal Income

* Change in personal income per capita over the past 10 years. 
* According to the correlation coefficient, changes in minimum wage and changes in personal income are positively correlated. 

GDP
===================================== 

## Column {data-width=550}

```{r}
ggplot(GDP, aes(x = X10change.y, y = pc10change)) +
  geom_point() + 
  geom_smooth() +
  xlab("Change in Minimum Wage ($)") +
  ylab("Change in Real GDP per Capita (Current $)") 
```

### Correlation Coefficient
```{r}
valueBox(0.4422169)
```

## Column {data-width=350}

### Gross Domestic Product

* Change in GDP per capita over the past 10 years. 
* GDP is the value added through production of goods and services. 
* GDP = private consumption + gross private investment + government investment + government spending + (exports – imports).
* Changes in minimum wage and changes in GDP are moderately correlated.  


